---
title: "LLeininger_KIPP_TakeHome"
author: "Lisa Leininger"
date: "5/19/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Summary:

```{r}
#simplify excel file names
SchoolGrade <- Copy_of_KTX_Analytics_Team_Performance_Task_Raw_Data
MAPscores <- MAP_scores
SpecialPrograms <- special_programs
```

```{r}
#load packages
library(plyr) #used for cleaning data
library(dplyr) # used for cleaning data
library(DataExplorer)
library(ggplot2)
library(RColorBrewer)
```

To clean this data set, my goal is to have all the information contained in one table with no redundancy. To accomplish this, I will first split up the MAPscores table into four separate tables characterized by test subject and semester taken. I will then join these four tables with the SchoolGrade and SpecialPrograms table, so that I have one dataframe where the entirety of data for each individual student is contained in one row. In the original data, the reading test was also coded two different ways, Reading and Read. I standardized this in excel before loading the data into RStudio.  I also made the assumption that if a student is coded as '.' or 'NULL' in Special Programs, then they are receiving general education services and I have recoded it as such using excel prior to loading the data into R.

```{r}
#Fall Math
fall_math_map <- MAPscores %>%
  filter(map_term == "Fall", subject =="Math")

#rename columns
fall_math_map <- plyr::rename(fall_math_map, c("rit_score"="fall_math_rit_score", "percentile"="fall_math_percentile", "quartile"="fall_math_quartile", "typical_fall_to_spring_growth"="math_typical_fall_to_spring_growth"))

#drop unneccessary columns
fall_math_map <- fall_math_map %>%
  select(student_id, year, fall_math_rit_score, fall_math_percentile, fall_math_quartile, math_typical_fall_to_spring_growth)


#Fall Reading
fall_read_map <- MAPscores %>%
  filter(map_term == "Fall", subject =="Read")

#rename columns
fall_read_map <- plyr::rename(fall_read_map, c("rit_score"="fall_read_rit_score", "percentile"="fall_read_percentile", "quartile"="fall_read_quartile", "typical_fall_to_spring_growth"="read_typical_fall_to_spring_growth"))

#drop unnecessary colums
fall_read_map <- fall_read_map %>%
  select(student_id, fall_read_rit_score, fall_read_percentile, fall_read_quartile, read_typical_fall_to_spring_growth)


#Spring Math
spring_math_map <- MAPscores %>%
  filter(map_term == "Spring", subject =="Math")

#rename columns
spring_math_map <- plyr::rename(spring_math_map, c("rit_score"="spring_math_rit_score", "percentile"="spring_math_percentile", "quartile"="spring_math_quartile"))

#drop unnecessary columns (including fall_to_spring_growth since it is not applicable in the spring)
spring_math_map <- spring_math_map %>%
  select(student_id, spring_math_rit_score, spring_math_percentile, spring_math_quartile)


#Spring Reading
spring_read_map <- MAPscores %>%
  filter(map_term == "Spring", subject =="Read")

#rename columns
spring_read_map <- plyr::rename(spring_read_map, c("rit_score"="spring_read_rit_score", "percentile"="spring_read_percentile", "quartile"="spring_read_quartile"))

#drop unnecessary columns (including fall_to_spring_growth since it is not applicable in the spring)
spring_read_map <- spring_read_map %>%
  select(student_id, spring_read_rit_score, spring_read_percentile, spring_read_quartile)
```

In the following section, I will join the four individual MAP score tables along with the school_grade and special_programs table into one table with all the relevant information.
```{r}
#inner join to combine all of the tables into one table with no redundancies
combined_data <- SchoolGrade %>%
  inner_join(SpecialPrograms, by = "student_id") %>%
  inner_join(fall_math_map, by = "student_id") %>%
  inner_join(fall_read_map, by = "student_id") %>%
  inner_join(spring_math_map, by = "student_id") %>%
  inner_join(spring_read_map, by = "student_id")

glimpse(combined_data)
```


```{r}
#convert necessary columns to numeric

#check classes
sapply(combined_data, class)

#convert to numeric
i <- c(3,6:19)
combined_data[ , i] <- apply(combined_data[ , i], 2,
                             function(x) as.numeric(as.character(x)))
#check conversion worked
sapply(combined_data, class)

```

Feature Engineering:
```{r}
#add raw growth columns for reading and math
combined_data2 <- combined_data %>%
  mutate(raw_fall_to_spring_math_growth = spring_math_rit_score - fall_math_rit_score) %>%
  mutate(raw_fall_to_spring_read_growth = spring_read_rit_score - fall_read_rit_score)

#add percent growth columns for reading and math
combined_data_3 <- combined_data2 %>%
  mutate(percentage_fall_to_spring_math_growth = raw_fall_to_spring_math_growth / fall_math_rit_score * 100) %>%
  mutate(percentage_fall_to_spring_read_growth = raw_fall_to_spring_read_growth / fall_read_rit_score * 100)

#add met growth math
combined_data_4 <- combined_data_3 %>%
  mutate(math_growth_met = ifelse(raw_fall_to_spring_math_growth >= math_typical_fall_to_spring_growth, "yes", "no"))

#add met growth readding
combined_data_final <- combined_data_4 %>%
  mutate(read_growth_met = ifelse(raw_fall_to_spring_read_growth >= read_typical_fall_to_spring_growth, "yes", "no"))
```


Check for missing values:
```{r}
plot_missing(combined_data_final, title = "Missing Data Across All Campuses")
```

```{r}
#look at age break down of schools
table(combined_data_final$school_name, combined_data_final$grade_level_2019)
```
Schools A, B, C, and D appear to be middle or intermediate. Schools E, F, and G appear to be elementary. I'm not sure what Grade 0 is but it is consistent between the three schools and so I'm going to leave it because I am not sure what to recode it as. I'm curious if there is a difference between schools and missing test scores, so I am going to check that out next.

```{r}
#School A
schoolA <- combined_data_final %>%
  filter(school_name == "School A")

plot_missing(schoolA, title = "Missing Data: School A")

#School B
schoolB <- combined_data_final %>%
  filter(school_name == "School B")

plot_missing(schoolB, title = "Missing Data: School B")

#School C
schoolC <- combined_data_final %>%
  filter(school_name == "School C")

plot_missing(schoolC, title = "Missing Data: School C")

#School D
schoolD <- combined_data_final %>%
  filter(school_name == "School D")

plot_missing(schoolD, title = "Missing Data: School D")

#School E
schoolE <- combined_data_final %>%
  filter(school_name == "School E")

plot_missing(schoolE, title = "Missing Data: School E")

#School F
schoolF <- combined_data_final %>%
  filter(school_name == "School F")

plot_missing(schoolF, title = "Missing Data: School F")

#School G
schoolG <- combined_data_final %>%
  filter(school_name == "School G")

plot_missing(schoolG, title = "Missing Data: School G")

```
School E seems to have done a really good job getting the majority of their students tested. Other schools have significantly more missing scores. It seems different kids are missing test scores from fall to spring and so we lose the opportunity to track their growth. School A is missing enough data to not be able to track growth for 23.78% of their students and school D is missing gorwth data for 35.97% of their students. The remaing schools are missing under 20% of growth data.


Now that I know which schools are elementary and middle, I am going to do further analysis on the subgroups separately.
```{r}
glimpse(combined_data_final)

```

8th Grade Math Analysis

```{r}
grade8 <- combined_data_final %>%
  filter(grade_level_2019 == 8)

table(grade8$school_name, grade8$program_name)

table(grade8$school_name, grade8$spring_math_quartile)

#compare expected growth to actual growth - this is tricky to read because the two need different scales
ggplot(grade8, aes(x=math_typical_fall_to_spring_growth, y=raw_fall_to_spring_math_growth, color=school_name, shape=program_name)) +geom_point()

summary(grade8)

#Fall score by School & Program
grade8 %>%
  ggplot( aes(x=school_name, y=fall_math_rit_score)) + 
  geom_rect(xmin = -Inf, ymin = 223, xmax = Inf, ymax = 237,
            fill = "darkolivegreen1")  +
  geom_boxplot() + 
  geom_jitter(width=0.2,alpha=0.8, aes(color=program_name)) +          labs(title = "Fall 8th Math Score by Campus", x = "Campus", y = "Math RIT Score") +
theme(plot.title = element_text(hjust=.5),axis.text.x = element_text( size = 12),axis.text.y = element_text(size = 12), legend.title = element_text(size = 12),
  legend.text = element_text(size=12)) 

#spring score by school & program
grade8 %>%
  ggplot( aes(x=school_name, y=spring_math_rit_score)) + 
  geom_rect(xmin = -Inf, ymin = 226, xmax = Inf, ymax = 241,
            fill = "darkolivegreen1")  +
  geom_boxplot() + 
  geom_jitter(width=0.2,alpha=0.8, aes(color=program_name)) +          labs(title = "Spring 8th Math Score by Campus", x = "Campus", y = "Math RIT Score") +
theme(plot.title = element_text(hjust=.5),axis.text.x = element_text( size = 12),axis.text.y = element_text(size = 12), legend.title = element_text(size = 12),
  legend.text = element_text(size=12))

#score by school and growth met
grade8 %>%
  ggplot( aes(x=school_name, y=spring_math_rit_score)) + 
  geom_rect(xmin = -Inf, ymin = 226, xmax = Inf, ymax = 241,
            fill = "darkolivegreen1")  +
  geom_boxplot() + 
  geom_jitter(width=0.2,alpha=0.8, aes(color=math_growth_met)) +labs(title = "Spring 8th Math Score by Campus", x = "Campus", y = "Math RIT Score") +
theme(plot.title = element_text(hjust=.5),axis.text.x = element_text( size = 12),axis.text.y = element_text(size = 12), legend.title = element_text(size = 12),legend.text = element_text(size=12))

#table of growth met and fall math quartile
table(grade8$math_growth_met, grade8$fall_math_quartile)

#table of growth met and spring math quartile
table(grade8$math_growth_met, grade8$spring_math_quartile)

#table of school and growth met
table(grade8$school_name, grade8$math_growth_met) 

#score by program name and growth met
grade8 %>%
  ggplot( aes(x=program_name, y=spring_math_rit_score)) + 
  geom_rect(xmin = -Inf, ymin = 226, xmax = Inf, ymax = 241,
            fill = "darkolivegreen1")  +
  geom_boxplot() + 
  geom_jitter(width=0.2,alpha=0.8, aes(color=math_growth_met)) +labs(title = "Spring 8th Math Score by Program Name", x = "Program Name", y = "Math RIT Score") +
theme(plot.title = element_text(hjust=.5),axis.text.x = element_text( size = 12),axis.text.y = element_text(size = 12), legend.title = element_text(size = 12),legend.text = element_text(size=12))

#table by program name and growth met
table(grade8$program_name, grade8$math_growth_met) 

#percent that met their goal
percent_math_met <- sum(grade8_complete$math_growth_met == "yes") / 277 * 100
percent_math_met

#percent that did not lose ground
percent_math_close <- sum(grade8_complete$raw_fall_to_spring_math_growth >=0) / 277 * 100
percent_math_close - percent_math_met

median(grade8_complete$math_typical_fall_to_spring_growth)

#percent that scored lower
percent_math_miss <- sum(grade8_complete$raw_fall_to_spring_math_growth <0) / 277 * 100
percent_math_miss
```

8th Grade Reading
```{r}
#Fall score by School & Program
grade8 %>%
  ggplot( aes(x=school_name, y=fall_read_rit_score)) + 
  geom_rect(xmin = -Inf, ymin = 214, xmax = Inf, ymax = 226,
            fill = "darkolivegreen1")  +
  geom_boxplot() + 
  geom_jitter(width=0.2,alpha=0.8, aes(color=program_name)) +          labs(title = "Fall 8th Grade Reading Scores by Campus", x = "Campus", y = "Reading RIT Score") +
theme(plot.title = element_text(hjust=.5),axis.text.x = element_text( size = 12),axis.text.y = element_text(size = 12), legend.title = element_text(size = 12),
  legend.text = element_text(size=12)) 

#spring score by school & program
grade8 %>%
  ggplot( aes(x=school_name, y=spring_read_rit_score)) + 
  geom_rect(xmin = -Inf, ymin = 217, xmax = Inf, ymax = 228,
            fill = "darkolivegreen1")  +
  geom_boxplot() + 
  geom_jitter(width=0.2,alpha=0.8, aes(color=program_name)) +          labs(title = "Spring 8th Grade Reading Scores by Campus", x = "Campus", y = "Reading RIT Score") +
theme(plot.title = element_text(hjust=.5),axis.text.x = element_text( size = 12),axis.text.y = element_text(size = 12), legend.title = element_text(size = 12),
  legend.text = element_text(size=12))

#score by school and growth met
grade8 %>%
  ggplot( aes(x=school_name, y=spring_read_rit_score)) + 
  geom_rect(xmin = -Inf, ymin = 217, xmax = Inf, ymax = 228,
            fill = "darkolivegreen1")  +
  geom_boxplot() + 
  geom_jitter(width=0.2,alpha=0.8, aes(color=read_growth_met)) +labs(title = "Spring 8th Grade Reading Scores by Campus", x = "Campus", y = "Reading RIT Score") +
theme(plot.title = element_text(hjust=.5),axis.text.x = element_text( size = 12),axis.text.y = element_text(size = 12), legend.title = element_text(size = 12),legend.text = element_text(size=12))

#table of growth met and fall reading quartile
table(grade8$read_growth_met, grade8$fall_read_quartile)

#table of growth met and spring reading quartile
table(grade8$read_growth_met, grade8$spring_read_quartile)

#table of school and growth met
table(grade8$school_name, grade8$read_growth_met) 

#score by program name and growth met
grade8 %>%
  ggplot( aes(x=program_name, y=spring_read_rit_score)) + 
  geom_rect(xmin = -Inf, ymin = 217, xmax = Inf, ymax = 228,
            fill = "darkolivegreen1")  +
  geom_boxplot() + 
  geom_jitter(width=0.2,alpha=0.8, aes(color=read_growth_met)) +labs(title = "Spring 8th Grade Reading Scores by Program Name", x = "Program Name", y = "Reading RIT Score") +
theme(plot.title = element_text(hjust=.5),axis.text.x = element_text( size = 12),axis.text.y = element_text(size = 12), legend.title = element_text(size = 12),legend.text = element_text(size=12))

#table by program name and growth met
table(grade8$program_name, grade8$read_growth_met)

#percent that met their goal
percent_read_met <- sum(grade8_complete$read_growth_met == "yes") / 277 * 100
percent_read_met

#percent that did not lose ground
percent_read_close <- sum(grade8_complete$raw_fall_to_spring_read_growth >=0) / 277 * 100
percent_read_close - percent_read_met

#percent that scored lower
percent_read_miss <- sum(grade8_complete$raw_fall_to_spring_read_growth <0) / 277 * 100
percent_read_miss

median(grade8_complete$read_typical_fall_to_spring_growth)
```



